quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space

### 2024-04-01 paper-quollr.Rmd
Add vis ref

---
title: "quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space"
abstract: >
  An abstract of less than 150 words.
draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      echo = FALSE)

@@

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(ggbeeswarm)
library(patchwork)

@@

<!--intro with S-curve with 5 methods-->

@@

<!--explain algorithm steps in summary-->

@@

```install.packages("quollr")```

@@

```devtools::install_github("JayaniLakshika/quollr")```

@@

@@

@@

The mains steps for the hexagonal binning algorithm can be executed by the main function `hex_binning()`, or can be run separately for more flexibility.
<!-- add about main function for hexagonal bining-->

@@

@@

@@

@@

@@

@@

## Tuning model parameters

@@

@@

@@

<!--2D model vis: triangular mesh, colored trimesh, removed edge trimesh-->
To visualize the 2D model, mainly three functions are used. As shown in Figure 1, `geom_trimesh()` to visualize the triangular mesh by adding a new layer to `ggplot()`. After identifying benchmark value to remove long edge, `vis_lg_mesh()` is used to visualize the triangular mesh by coloring the small and long edges. As shown in Figure 1, the small and long edges are colored by black and red respectively. Following this, `vis_rmlg_mesh()` is used to visualize smoothed 2D model which is the 2D model after removing the long edges (see Figure 1). In `vis_lg_mesh()` and `vis_rmlg_mesh()`, `benchmark_value` argument controls the edge removal in 2D.

```{r, out.width="100%", fig.cap="Visualization of model generated for UMAP embedding for S-curve data in 2D space: (a) triangular mesh, (b), triangular mesh colored by edge type, and (c) triangular mesh after removing the long edges. The edges which has the length greater than $0.5$ are assigned as long edges. To obtain smooth representation in 2D space, the long edges are removed."}

@@

<!--need to add screenshots of the frames and add the link of the video-->
```{r}
tour
```

@@

@@

@@

```{r}
library(Rtsne)

tSNE_fit <- clust_df |>
  dplyr::select(-ID) |>
  dplyr::select(where(is.numeric)) |>
  Rtsne::Rtsne(perplexity = 39, pca = FALSE, pca_center = FALSE, normalize = FALSE, dims = 2)

@@

tSNE_df_plot <- tSNE_df |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.5, colour="#e41a1c", size = 0.5) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.text = element_text(size = 5),
        axis.title = element_text(size = 7))

tSNE_df_plot
```

@@

```{r}
mse_clust_plot <- ggplot(mse_df_clust, aes(x = num_bins,
                                       y = log(mse)
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  ylab("MSE") +
  xlab("total number of bins")

mse_clust_plot
```

>>>>

---
title: "quollr: An R Package for Visualizing 2D Models from Nonlinear Dimension Reductions in High Dimensional Space"
abstract: >
  Non-Linear Dimension Reduction (NLDR) techniques have emerged as powerful tools to visualize high-dimensional ($p$-D) data. However, their complexity and parameter choices may lead to distrustful or misleading results. The R package \CRANpkg{quollr} is developed as a new tool to help to determine which method, which parameter choice provide the most useful representation of high-D data. Clustering data from \CRANpkg{cardinalR} package, is used to illustrate the algorithm and its use within the paackage.
draft: true
author:  
  - name: Jayani P.G. Lakshika
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: https://jayanilakshika.netlify.app/
    orcid: 0000-0002-6265-6481
    email:  \email{jayani.piyadigamage@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    url: http://www.dicook.org/
    email: dicook@monash.edu
    orcid: 0000-0002-3813-7155
  - name: Paul Harrison
    affiliation: Monash University
    address: MGBP, BDInstitute, VIC 3800 Australia
    email: paul.harrison@monash.edu
    orcid: 0000-0002-3980-268X
  - name: Michael Lydeamore
    affiliation: Monash University
    address: Department of Econometrics and Business Statistics, VIC 3800 Australia
    email: michael.lydeamore@monash.edu
    orcid: 0000-0001-6515-827X
  - name: Thiyanga S. Talagala
    affiliation: University of Sri Jayewardenepura
    address: Department of Statistics, Gangodawila, Nugegoda 10100 Sri Lanka
    url: https://thiyanga.netlify.app/
    email: ttalagala@sjp.ac.lk
    orcid: 0000-0002-0656-9789
type: package
creative_commons: CC BY
date: "`r Sys.Date()`"
preamble: >
  \usepackage{amsmath}
  \usepackage{array}
output: 
 rjtools::rjournal_web_article:
    css: "style.css"
    keep_md: true
bibliography: RJreferences.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  cache=FALSE, 
  message=FALSE, 
  warning=FALSE)

@@

```{r load-libraries}
library(quollr)
library(tibble)
library(knitr)
library(kableExtra)
library(ggplot2)
library(dplyr)
library(ggbeeswarm)
library(patchwork)
library(Rtsne)

@@

<!-- research gap: add about hexbin pkg, and emphasize that in our package provide regular hexagons-->
<!-- objective: introduce a new tool to help to determine which method, which parameter choice provide the most useful representation of high-D data.--> 
<!--intro with S-curve with 5 methods-->

@@

<!--explain algorithm steps in summary-->
Our algorithm mainly consists of two steps: (1) construct the model in 2D space, and (2) lift the model into high dimensions. Constructing the model in 2D space involves: (1) scaling NLDR data, (2) binning data, (3) obtain bin centroids/ means, and (4) triangulate bin centroids/ means. The implementation of these steps are described in detail in the following section.

@@

```r
install.packages("quollr")
```

@@

```r
devtools::install_github("JayaniLakshika/quollr")
```

@@

```{r, echo=TRUE, eval=FALSE}
model_object <- fit_highd_model( training_data = s_curve_noise_training, 
                                 nldr_df_with_id = s_curve_noise_umap_scaled, 
                                 x = "UMAP1", y = "UMAP2", 
                                 num_bins_x = num_bins_x_s_curve, 
                                 num_bins_y = num_bins_y_s_curve, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_s_curve,
                                 is_bin_centroid = TRUE,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "UMAP", 
                                 col_start_highd = "x")
```

@@

```{r, echo=TRUE, eval=FALSE}
s_curve_noise_umap_scaled <- gen_scaled_data(data = s_curve_noise_umap, x = "UMAP1", y = "UMAP2")
```

@@

```{r, echo=TRUE, eval=FALSE}
bin_list <- calc_bins(data = s_curve_noise_umap_scaled, x = "UMAP1", 
                      y = "UMAP2", hex_size = NA, buffer_x = NA, buffer_y = NA)

num_bins_x <- bin_list$num_x
num_bins_y <- bin_list$num_y
```

The mains steps for the hexagonal binning algorithm can be executed by the main function `hex_binning()`, or can be run separately for more flexibility.
<!-- add about main function for hexagonal binning-->

```{r, echo=TRUE, eval=FALSE}
hb_obj <- hex_binning(data = s_curve_noise_umap_scaled, x = "UMAP1", 
                      y = "UMAP2", num_bins_x = num_bins_x, 
                      num_bins_y = num_bins_y, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = NA, 
                      col_start = "UMAP")

@@

```{r, echo=TRUE, eval=FALSE}
centroid_list <- gen_centroids(data = s_curve_noise_umap_scaled, x = "UMAP1", 
                               y = "UMAP2", num_bins_x = num_bins_x,
                               num_bins_y = num_bins_y, x_start = NA,
                               y_start = NA, buffer_x = NA, buffer_y = NA, 
                               hex_size = NA)

all_centroids_df <- as.data.frame(do.call(cbind, centroid_list))
```

@@

```{r, echo=TRUE, eval=FALSE}
all_hex_coordinates_list <- gen_hex_coord(centroids_df = all_centroids_df,
                                            hex_size = NA)
```

@@

```{r, echo=TRUE, eval=FALSE}
df_with_hb_id_list <- assign_data(data = s_curve_noise_umap_scaled |> 
                                    dplyr::select(-ID), 
                                  centroid_df = all_centroids_df, 
                                  col_start = "UMAP")
umap_data_with_hb_id <- as.data.frame(do.call(cbind, df_with_hb_id_list))
```

```{r, echo=TRUE, eval=FALSE}
std_counts_list <- compute_std_counts(data_hex_id = df_with_hex_id)
counts_df <- as.data.frame(do.call(cbind, std_counts_list))
```

@@

```{r, echo=TRUE, eval=FALSE}
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
```

```{r, echo=TRUE, eval=FALSE}
df_bin_means <- extract_hexbin_mean(nldr_df_with_hex_id = umap_with_hb_id,
                                        counts_df = counts_df)
```

@@

```{r, echo=TRUE, eval=FALSE}
tr1_object <- tri_bin_centroids(hex_df = df_bin_centroids, x = "c_x", y = "c_y")
tr_from_to_df <- gen_edges(tri_object = tr1_object)
```

@@

```{r, echo=TRUE, eval=FALSE}
df_all <- dplyr::bind_cols(s_curve_noise_training |> dplyr::select(-ID),
                           umap_data_with_hb_id)

df_bin <- avg_highd_data(data = df_all, col_start = "x")
```

## Model parameters

@@

```{r, echo=TRUE, eval=FALSE}
df_bin_centroids_low <- df_bin_centroids |> dplyr::filter(std_counts <= 0.43)

low_dens_hex <- find_low_dens_hex(df_bin_centroids_all = df_bin_centroids, 
                  num_bins_x = num_bins_x, 
                  df_bin_centroids_low = df_bin_centroids_low)
```

```{r, echo=TRUE, eval=FALSE}
distance_df <- cal_2d_dist(tr_coord_df = tr_from_to_df, start_x = "x_from", 
                           start_y = "y_from", end_x = "x_to", end_y = "y_to",
                           select_vars = c("from", "to", "distance"))

benchmark_val <- find_lg_benchmark(distance_edges = distance_df, distance_col = "distance")
```

@@

```{r, echo=TRUE}
gen_summary(test_data = s_curve_noise_training, prediction_df = pred_df_training,
df_bin = df_bin_s_curve, col_start = "x")
```

```{r, echo=TRUE}
gen_summary(test_data = s_curve_noise_test, prediction_df = pred_df_test,
df_bin = df_bin_s_curve, col_start = "x")
```

@@

<!--2D model vis: triangular mesh, colored trimesh, removed edge trimesh-->
To visualize the 2D model, mainly three functions are used. As shown in Figure \@ref(fig:mesh-plots)a, `geom_trimesh()` to visualize the triangular mesh by adding a new layer to `ggplot()`. After identifying benchmark value to remove long edge, `vis_lg_mesh()` is used to visualize the triangular mesh by coloring the small and long edges. As shown in Figure \@ref(fig:mesh-plots)b, the small and long edges are colored by black and red respectively. Following this, `vis_rmlg_mesh()` is used to visualize smoothed 2D model which is the 2D model after removing the long edges (see Figure \@ref(fig:mesh-plots)c). In `vis_lg_mesh()` and `vis_rmlg_mesh()`, `benchmark_value` argument controls the edge removal in 2D.

```{r mesh-plots, out.width="100%", fig.cap="Visualization of model generated for UMAP embedding for S-curve data in 2D space: (a) triangular mesh, (b), triangular mesh colored by edge type, and (c) triangular mesh after removing the long edges. The edges which has the length greater than $0.5$ are assigned as long edges. To obtain smooth representation in 2D space, the long edges are removed."}

@@

```{r, echo=TRUE, eval=FALSE}
tour <- show_langevitour(df = df_all, df_b = df_bin, 
                 df_b_with_center_data = df_bin_centroids, benchmark_value = 0.5, 
                 distance = distance_df, distance_col = "distance", 
                 use_default_benchmark_val = FALSE, col_start = "x")
```

<!--need to add screenshots of the frames and add the link of the video-->
```{r tour-s-curve}
tour
```

@@

We illustrate the use of the algorithm by applying it to `clusters_different_shapes_diff_num_points` dataset in `cardinalR` package. The tSNE is used in the example for the demonstration (see Figure \@ref(fig:tsne-clust)).

@@

<!--need to add screen shots of the original data-->

@@

```{r}
tSNE_fit <- clust_df |>
  dplyr::select(-ID) |>
  dplyr::select(where(is.numeric)) |>
  Rtsne::Rtsne(perplexity = 39, pca = FALSE, pca_center = FALSE, normalize = FALSE, dims = 2)

@@

tSNE_df_plot <- tSNE_df |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.5, colour="#e41a1c", size = 0.5) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        axis.text = element_text(size = 5),
        axis.title = element_text(size = 7))
```

```{r tsne-clust, out.width="100%", fig.cap="tSNE representation of data with perplexity: 39. Is this representation accurately capture the structure in high dimensions?"}
tSNE_df_plot
```

## Selecting parameter values for hexagonal binning

We previously introduced three key parameters, *hex_size*, *x_start*, and *y_start*, in our explanation of the algorithm. While their best choice is unknown, a visualization tool enables their tuning such that user can select reasonable values.

Consider the relationships between *total number of bins*, *starting point*, and MSE. Increase of *total number of bins* usually reduces the MSE.

<!--hex_size, starting point (x_start, y_start)-->
<!--talk about to use bin centroids or bin means-->

@@

```{r}
mse_clust_plot <- ggplot(mse_df_clust, aes(x = num_bins,
                                       y = log(mse)
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 588, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  ylab("log(MSE)") +
  xlab("total number of bins")
```

```{r, fig.cap="Parameter tuning plots, where the best choice of parameter at a large drop in MSE. Here, these are at *total number of bins* = 180."}
mse_clust_plot
```

## Selecting parameter values for the model

<!--benchmark value to remove low density hexagons, long edges-->

## Fit the model

To perform the algorithm on the clustering data, we first scale the tSNE data. We then let *number of bins along the x-axis* be $21$, *number of bins along the y-axis* be $28$, and *hex_size* be 0.03; see [Selecting parameter values] for further discussion and guidance regarding selection of these choices.

### 2024-04-01 paper-quollr.tex
Add vis ref

\abstract{%
An abstract of less than 150 words.
}

@@

\hypertarget{tuning-model-parameters}{%
\subsection{Tuning model parameters}\label{tuning-model-parameters}}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-4-1} \caption{Predicted 2D embedding overlaid with S-curve training UMAP data: (a) Prediction of 2D embeddings for S-curve training data, and (b) Prediction of 2D embeddings for S-curve test data. Once the high-D model is generated, first, for the new data points the nearest high-D mappings of hexagonal bin centroids find by computing Euclidean distance. Then, mapping back the 2D hexagonal bin centroids gives the prediction for the new data points.}\label{fig:unnamed-chunk-4}
\end{figure}

@@

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-5-1} \caption{Visualization of model generated for UMAP embedding for S-curve data in 2D space: (a) triangular mesh, (b), triangular mesh colored by edge type, and (c) triangular mesh after removing the long edges. The edges which has the length greater than $0.5$ are assigned as long edges. To obtain smooth representation in 2D space, the long edges are removed.}\label{fig:unnamed-chunk-5}
\end{figure}

@@

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-6-1.pdf}

@@

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-8-1.pdf}

@@

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-11-1.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-12-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-12-2.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-13-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-13-2.pdf}

>>>>

\abstract{%
Non-Linear Dimension Reduction (NLDR) techniques have emerged as powerful tools to visualize high-dimensional (\(p\)-D) data. However, their complexity and parameter choices may lead to distrustful or misleading results. The R package \CRANpkg{quollr} is developed as a new tool to help to determine which method, which parameter choice provide the most useful representation of high-D data. Clustering data from \CRANpkg{cardinalR} package, is used to illustrate the algorithm and its use within the paackage.
}

@@

\hypertarget{model-parameters}{%
\subsection{Model parameters}\label{model-parameters}}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-4-1} \caption{Predicted 2D embedding overlaid with UMAP embedding of S-curve training data: (a) prediction of 2D embeddings for S-curve training data, and (b) prediction of 2D embeddings for S-curve test data. Once the high-D model is generated, first, for the new data points the nearest high-D mappings of hexagonal bin centroids find by computing Euclidean distance. Then, mapping back the 2D hexagonal bin centroids gives the prediction for the new data points.}\label{fig:unnamed-chunk-4}
\end{figure}

@@

\begin{verbatim}
gen_summary(test_data = s_curve_noise_training, prediction_df = pred_df_training,
df_bin = df_bin_s_curve, col_start = "x")
\end{verbatim}

\begin{verbatim}
#> $error
#> [1] 35.79497
#> 
#> $mse
#> [1] 0.1114669
#> 
#> $aic
#> [1] -815.8643
\end{verbatim}

\begin{verbatim}
gen_summary(test_data = s_curve_noise_test, prediction_df = pred_df_test,
df_bin = df_bin_s_curve, col_start = "x")
\end{verbatim}

\begin{verbatim}
#> $error
#> [1] 15.4298
#> 
#> $mse
#> [1] 0.1546555
#> 
#> $aic
#> [1] 9.352806
\end{verbatim}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/unnamed-chunk-7-1} \caption{Visualization of model generated for UMAP embedding for S-curve data in 2D space: (a) triangular mesh, (b), triangular mesh colored by edge type, and (c) triangular mesh after removing the long edges. The edges which has the length greater than $0.5$ are assigned as long edges. To obtain smooth representation in 2D space, the long edges are removed.}\label{fig:unnamed-chunk-7}
\end{figure}

@@

\includegraphics{paper-quollr_files/figure-latex/tour-s-curve-1.pdf}

@@

We illustrate the use of the algorithm by applying it to \texttt{clusters\_different\_shapes\_diff\_num\_points} dataset in \texttt{cardinalR} package. The tSNE is used in the example for the demonstration (see Figure \ref{fig:tsne-clust}).

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper-quollr_files/figure-latex/tsne-clust-1} \caption{tSNE representation of data with perplexity: 39. Is this representation accurately capture the structure in high dimensions?}\label{fig:tsne-clust}
\end{figure}

\hypertarget{selecting-parameter-values-for-hexagonal-binning}{%
\subsection{Selecting parameter values for hexagonal binning}\label{selecting-parameter-values-for-hexagonal-binning}}

We previously introduced three key parameters, \emph{hex\_size}, \emph{x\_start}, and \emph{y\_start}, in our explanation of the algorithm. While their best choice is unknown, a visualization tool enables their tuning such that user can select reasonable values.

Consider the relationships between \emph{total number of bins}, \emph{starting point}, and MSE. Increase of \emph{total number of bins} usually reduces the MSE.

\begin{figure}
\centering
\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-13-1.pdf}
\caption{\label{fig:unnamed-chunk-13}Parameter tuning plots, where the best choice of parameter at a large drop in MSE. Here, these are at \emph{total number of bins} = 180.}
\end{figure}

\hypertarget{selecting-parameter-values-for-the-model}{%
\subsection{Selecting parameter values for the model}\label{selecting-parameter-values-for-the-model}}

\hypertarget{fit-the-model}{%
\subsection{Fit the model}\label{fit-the-model}}

To perform the algorithm on the clustering data, we first scale the tSNE data. We then let \emph{number of bins along the x-axis} be \(21\), \emph{number of bins along the y-axis} be \(28\), and \emph{hex\_size} be 0.03; see {[}Selecting parameter values{]} for further discussion and guidance regarding selection of these choices.

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-14-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-14-2.pdf}

\includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-15-1.pdf} \includegraphics{paper-quollr_files/figure-latex/unnamed-chunk-15-2.pdf}
